Locality and Computational Reliability of Linear Response Calculations for Molecular Systems

### 2018-07-12 AnalyticStructure_main.tex
update

%\title{On the analytic structure of the linear response susceptibility in open systems}
\title{Locality and Computational Reliability of Linear Response Calculations for Molecular Systems}% Localization properties, observable character and computational reliability of the linear response in open systems}
\author{Marco D'Alessandro}%
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}
\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), INAC, CEA-UGA, 17 Av. des Martyrs,
38054 Grenoble, France}
\date{\today}

@@

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed 
for ground state calculations of molecules. 
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behaviour of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will also reveal useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in the Appendix.

@@

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the 
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed hamiltonian can be extracted with traditional GS techniques, 
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
%\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
% is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
% solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line) 
of $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated to the largest 
box.}
\end{figure}

@@

It is therefore much more interesting to concentrate on LR quantities that are defined independently from the perturbing operators and target observables.
This can be achieved by studying the 
\emph{susceptibility functional}, defined implicitly trough the operator equation  
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \left(\omega -\mathcal L \right)^{-1} \commutator{\op \Phi(\omega)}{\dmnot} = \sop{{\cal \chi}}(\omega) \op\Phi(\omega) \;. 
\ee
%This alternative formulation of the problem has the advantage of providing general information on the nature of the excitation spectrum. Results achieved in this way thus describes an intrinsic 
%feature of the system under inspection and determine the properties of the LR independently from the specific choice of the perturbing field. 
The susceptibility (super)operator $\cal \chi(\omega)$ is therefore closely related to the resolvent of the Liouvillian of Eq.~\eqref{LiouvillianRhopomegaDef1}, which is completely defined 
from the molecule hamiltonian $\hnot[\dm]$.
The ``excitation modes''  of the molecule are defined through the \emph{excitation operators}, satisfying
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$. 
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations defined in that way may be considered as a basis of the operators that satisfy the transverse condition as in  
Eq.~\eqref{RhopTransverseDef1}, and enable us to express the susceptibility as a spectral decomposition:
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) \cdot   = \sum_{\{a\}} \op E_{a} \,
\frac{\trace{\commutator{\op{\tilde E}_a}{\dmnot}\cdot}}{\omega-\Omega_a} \;.
\ee

@@

With the same spirit of the previous sections, we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
\ket{\phi^a_p} &= \op G_H(\Omega_a+\eps_p)\left(\op V \ket{\phi^a_p} + \ket{s_p[\op E_a]} \right) \;, \nn \\
\bra {\chi^a_p} &=\left(\bra{\chi^a_p} \op V  + \bra{s_p[\op E_a]}\right)  \op G_H(-\Omega_a+\eps_p)\;,
\end{align}
where the source terms are here defined as:
\be
 \ket{s_p[\op O]} =  \op Q_0 \op V'[\op O] \ket{\psi_p}\,, \quad
 \bra{s_p[\op O]} =   \bra{\psi_p} \op V'[\op O] \op Q_0 \nn \;.
\ee
We first point out that the Helmholtz kernel associated to the $\ket{\chi^a_p}$ state 
contains an exponential damping factor for \emph{all} the positive values of $\Omega_a$.
The asymptotic long-range behaviour of $\ket{\chi^a_p}$ thus depends only on the locality properties of 
$\bra{s_p[\op E_a]}$, which in turn is related to the behavior of the operator $\op V'[\op E_a]$.
% As in the previous section, we assume the hypothesis that such operator is such that
% the state $\op Q_0 \op V'[\op E^t_a] \ket{\psi_p}$ is localized, at least when \emph{both}
% the states $\phi_p^a$ $\chi_p^a$ have bound-state-like behaviour.

On the other hand, the long-range behaviour of $\op G_H(\Omega_a+\eps_p)$ depends from 
$\Omega_a$, with a threshold of value $\Omega_a=|\eps_p|$. 
For energies above this value, the behaviour of the state $\ket{\phi^a_p}$ is 
delocalized regardless of the particular nature of the operators $\op V$ and $\op V'$.
This means that, if an excitation operator $\op E_a$ has an energy such that
$\Omega_a>|\eps_p|$  for at least one of the $p \in a$ defining the excitation,
such operator \emph{cannot} be parametrized by employing localized states only.

@@

In Fig.~\ref{CO_exc} and \ref{C6H6_exc} we plot the DOS of the Liouvillian for the $CO$  and Benzene molecules, 
for two different numerical setups of increasing size of the simulation domain. 
The figures show clearly that all the excitations of energy $\Omega_a$ lying below the 
molecule ionization potential (IP) have a clear localized behaviour, as their energy does not vary with the size of the simulation domain.
On the other hand, above the first ionization threshold the excitation energies have an energy that strongly depends on the computational
setup, which is in line with the expected pseudo-continuum character of this sector. 
In addition, for the benzene molecule it is also possible to 
identify excitations of energy above IP that still possess a localized behaviour, fulfilling the condition $\Omega_a < |\eps_a|$.
In Fig.~\ref{c6h6_excLand} we collect the excitations of the benzene molecule following this criterion:
the agreement with the expected analytic structure presented in Fig.~\ref{ExcitationLandscape} is remarkable.

@@

On the other hand, the excitation energies which lie above these thresholds are only pseudo-poles of $\chi(\omega)$, 
and will \emph{strongly} depend on the computational setup. This is visible for energies higher than
IP, and it has been pointed out several times (see e.g. \cite{giustino2012,giustino2014}) in the literature.
These excitations belong to a continuum sector that embeds the above mentioned discrete excitations.
We remark that we are not referring here to the continuum eigenstates of the 
unperturbed GS hamiltonian, %which is not in itself important for expressing the linear susceptivity, 
but to the continuum of excitations expressed as eigenstate of the 
Liouvillian superoperator of Eq.~\eqref{LiouZeroDef1}, needed to 
express the spectral representation of the linear susceptibility.

Both the sectors of the excitations energies have to be considered in view of the extraction of computationally reliable quantities, and this \emph{even below} the first IP.
For generic LR quantities, which are functionals of the linear susceptivity, 
the quality of the $\chi(\omega)$ (super)operator will depend
on the frequency $\omega$ of interest: a given spectral representation may provide high-quality results for static $\omega=0$ regimes, while being
unable to provide convergence for high-energy absorption spectra.
On the contrary, absorption spectra below IP are relatively easy to converge, as they are observables which depend only on localized quantities.

@@

\subsection{$CO$}
The molecule is oriented along the $z$ axis and the $z$-dimension of the box is equal to 11.5, 15.1 and 16.6 \AA.
280 unoccupied states have been computed in each setup, which span an energy range of the empty KS orbitals of 30.7, 19.3 and 15.4 eV respectively. 
Such energy values should not be confused with the excitations energy ranges reported in Figs.~\ref{co_spectrum} and \ref{CO_exc}. 
Once again, such choices of unoccupied states ensures convergence in of excitations spectum in the presented range.

>>>>

\title{Locality and Computational Reliability of Linear Response Calculations for Molecular Systems}% Localization properties, observable character and computational reliability of the linear response in open systems}
\author{Marco D'Alessandro}%
\affiliation{Istituto di Struttura della Materia-CNR (ISM-CNR), Via del Fosso del Cavaliere 100, 00133 Roma, Italia}
\author{Luigi Genovese}
\affiliation{Laboratoire de Simulation Atomistique (LSim), INAC, CEA-UGA, 17 Av. des Martyrs,
38054 Grenoble, France}
\date{\today}

@@

The paper is organized as follows.
Initially we will inspect how the question of reproducibility may be expressed 
for ground state calculations of molecules. 
Following these guidelines, we will then move to LR equations,
by first considering the equations of motion of the response density operator, given
the application of a perturbing field.
Then we will consider the behavior of the ``free oscillations'', i.e. the excited states
of the molecule.
Our considerations will also reveal useful for the understanding of the \emph{analytic structure} of LR quantities like
the linear susceptibility of the system (i.e. the reducible polarizability).
We support our considerations with numerical results on simple LR-TDDFT quantities.
For readability purposes, the details of each calculation are explained in the Appendix.

@@

Such locality property of the fluctuation state can be directly confirmed for static perturbations at $\omega=0$.
Let us consider a static perturbing operator $\op\Phi_s$ that can be restricted to a finite domain. 
Any local operator, like \emph{e.g.} a static electric field $\mathbf F$, with associated scalar potential $\Phi_s[\mathbf F](\r) = -\mathbf F \cdot \r$, would satisfy such property.
We consider the perturbed Hamiltonian $\hnot[\dm_{\Phi_s}] + \op\Phi_s[\mathbf F]$, where $\dm_{\Phi_s}$ represents the GS density of the perturbed system, and assume that the 
perturbation strength ($|\mathbf F|$ in the above example) is sufficiently small to be consistent with a LR treatment of the problem. 
In this case, the occupied eigenfunction $\ket{\psi_p^{\Phi_s}}$ of the perturbed Hamiltonian can be extracted with traditional GS techniques, 
as from the discussion around Eq.~\eqref{SEHelmholtz1}, they exhibit bound-state behavior.
%\footnote{The fulfilling of Eq.~\eqref{SEHelmholtz1} for energies below threshold
% is not \textit{per se} a guarantee of a localized $\ket{\psi}$. With such equation we demonstrate that, \emph{in the case when} a localized
% solution exists, this cannot be found for energies above threshold.}
\begin{figure}[t]
\includegraphics[scale=0.68]{CO_statPolvsBox.eps}
\caption{\label{co_alphaStatic}(Color online) Convergence of the statical polarizability $\alpha_{zz}$ (blue continuous line) and of the ground state dipole $d_z$ (red dotted line) 
of $CO$ molecule as a function of the size of the computational domain. The inset evidences the relative convergence (in percent) of the two quantities w.r.t. the corresponding value associated to the largest 
box.}
\end{figure}

@@

It is therefore much more interesting to concentrate on LR quantities that are defined independently from the perturbing operators and target observables.
This can be achieved by studying the 
\emph{susceptibility functional}, defined implicitly trough the operator equation  
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \left(\omega -\mathcal L \right)^{-1} \commutator{\op \Phi(\omega)}{\dmnot} = \sop{{\cal \chi}}(\omega) \op\Phi(\omega) \;. 
\ee
%This alternative formulation of the problem has the advantage of providing general information on the nature of the excitation spectrum. Results achieved in this way thus describes an intrinsic 
%feature of the system under inspection and determine the properties of the LR independently from the specific choice of the perturbing field. 
The susceptibility (super)operator $\cal \chi(\omega)$ is therefore closely related to the resolvent of the Liouvillian of Eq.~\eqref{LiouvillianRhopomegaDef1}, which is completely defined 
from the molecule Hamiltonian $\hnot[\dm]$.
The ``excitation modes''  of the molecule are defined through the \emph{excitation operators}, satisfying
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E_a = \Omega_a \op E_a \qq \op{\tilde E}_a \Liouv = \Omega_a \op{\tilde E}_a \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$. 
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}_a\op E_b} = \delta_{ab} \;.
\ee
Excitations defined in that way may be considered as a basis of the operators that satisfy the transverse condition as in  
Eq.~\eqref{RhopTransverseDef1}, and enable us to express the susceptibility as a spectral decomposition:
\be\lb{RhopExcitationDef1}
\sop{{\cal \chi}}(\omega) \cdot   = \sum_{\{a\}} \op E_{a} \,
\frac{\trace{\commutator{\op{\tilde E}_a}{\dmnot}\cdot}}{\omega-\Omega_a} \;.
\ee

@@

With the same spirit of the previous sections, we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
\ket{\phi^a_p} &= \op G_H(\Omega_a+\eps_p)\left(\op V \ket{\phi^a_p} + \ket{s_p[\op E_a]} \right) \;, \nn \\
\bra {\chi^a_p} &=\left(\bra{\chi^a_p} \op V  + \bra{s_p[\op E_a]}\right)  \op G_H(-\Omega_a+\eps_p)\;,
\end{align}
where the source terms are here defined as:
\be
 \ket{s_p[\op O]} =  \op Q_0 \op V'[\op O] \ket{\psi_p}\,, \quad
 \bra{s_p[\op O]} =   \bra{\psi_p} \op V'[\op O] \op Q_0 \nn \;.
\ee
We first point out that the Helmholtz kernel associated to the $\ket{\chi^a_p}$ state 
contains an exponential damping factor for \emph{all} the positive values of $\Omega_a$.
The asymptotic long-range behavior of $\ket{\chi^a_p}$ thus depends only on the locality properties of 
$\bra{s_p[\op E_a]}$, which in turn is related to the behavior of the operator $\op V'[\op E_a]$.
% As in the previous section, we assume the hypothesis that such operator is such that
% the state $\op Q_0 \op V'[\op E^t_a] \ket{\psi_p}$ is localized, at least when \emph{both}
% the states $\phi_p^a$ $\chi_p^a$ have bound-state-like behaviour.

On the other hand, the long-range behavior of $\op G_H(\Omega_a+\eps_p)$ depends from 
$\Omega_a$, with a threshold of value $\Omega_a=|\eps_p|$. 
For energies above this value, the behavior of the state $\ket{\phi^a_p}$ is 
delocalized regardless of the particular nature of the operators $\op V$ and $\op V'$.
This means that, if an excitation operator $\op E_a$ has an energy such that
$\Omega_a>|\eps_p|$  for at least one of the $p \in a$ defining the excitation,
such operator \emph{cannot} be parametrized by employing localized states only.

@@

In Fig.~\ref{CO_exc} and \ref{C6H6_exc} we plot the DOS of the Liouvillian for the $CO$  and Benzene molecules, 
for two different numerical setups of increasing size of the simulation domain. 
The figures show clearly that all the excitations of energy $\Omega_a$ lying below the 
molecule ionization potential (IP) have a clear localized behavior, as their energy does not vary with the size of the simulation domain.
On the other hand, above the first ionization threshold the excitation energies have an energy that strongly depends on the computational
setup, which is in line with the expected pseudo-continuum character of this sector. 
In addition, for the benzene molecule it is also possible to 
identify excitations of energy above IP that still possess a localized behavior, fulfilling the condition $\Omega_a < |\eps_a|$.
In Fig.~\ref{c6h6_excLand} we collect the excitations of the benzene molecule following this criterion:
the agreement with the expected analytic structure presented in Fig.~\ref{ExcitationLandscape} is remarkable.

@@

On the other hand, the excitation energies which lie above these thresholds are only pseudo-poles of $\chi(\omega)$, 
and will \emph{strongly} depend on the computational setup. This is visible for energies higher than
IP, and it has been pointed out several times (see e.g. \cite{giustino2012,giustino2014}) in the literature.
These excitations belong to a continuum sector that embeds the above mentioned discrete excitations.
We remark that we are not referring here to the continuum eigenstates of the 
unperturbed GS Hamiltonian, %which is not in itself important for expressing the linear susceptivity, 
but to the continuum of excitations expressed as eigenstate of the 
Liouvillian superoperator of Eq.~\eqref{LiouZeroDef1}, needed to 
express the spectral representation of the linear susceptibility.

Both the sectors of the excitations energies have to be considered in view of the extraction of computationally reliable quantities, and this \emph{even below} the first IP.
For generic LR quantities, which are functionals of the linear susceptibility, 
the quality of the $\chi(\omega)$ (super)operator will depend
on the frequency $\omega$ of interest: a given spectral representation may provide high-quality results for static $\omega=0$ regimes, while being
unable to provide convergence for high-energy absorption spectra.
On the contrary, absorption spectra below IP are relatively easy to converge, as they are observables which depend only on localized quantities.

@@

\subsection{$CO$}
The molecule is oriented along the $z$ axis and the $z$-dimension of the box is equal to 11.5, 15.1 and 16.6 \AA.
280 unoccupied states have been computed in each setup, which span an energy range of the empty KS orbitals of 30.7, 19.3 and 15.4 eV respectively. 
Such energy values should not be confused with the excitations energy ranges reported in Figs.~\ref{co_spectrum} and \ref{CO_exc}. 
Once again, such choices of unoccupied states ensures convergence in of excitations spectrum in the presented range.
